Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015

### 2018-01-23 paper/paper.tex
Last commit before delete local folder - everything moved to thesis/ repo

% Things to remember:
% The working directory is not where the data is.
% Change working directory to "/Users/cimentadaj/Downloads/gitrepo/Achievement_Gap_PISA/paper"
% for everything below to work.

% There's a code chunk named modeling which lasts about 3-4 hours to run. I explain how to get
% around that in a comment above the code chunk. For down there.

\documentclass[11pt, a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\bibliographystyle{apalike}
\pagestyle{headings}

\usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
 }
\usepackage{pdflscape}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{titling}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

\title{Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015}
\author{Jorge Cimentada}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\setlength{\parindent}{2em}
\setlength{\parskip}{1em}
\showboxdepth=5
\showboxbreadth=5

\maketitle

\begin{abstract}
The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has uncovered a wide variety of results. The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year. Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce. This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective. We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate. We find that this gap is strongly correlated with a country's overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%. We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the aversive effects of tracking.
\end{abstract}

% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
% code chunk

% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
% compiling sweave, comment the test_diff expressions and read as rds.

% Remember to uncomment when compiling sweave again, so the chunk remains intact.

\section{Introduction}

The impact of inequality, be it economic or of non pecuniary causes, is a crucial determinant for the opportunities of a child. Evidence from disciplines as disparate as epidemiology, psychology, sociology and economics keep piling evidence that suggests that inequality in the life course is something present in virtually all societies. This is something that politicians and policymakers are aware of and, yet despite their many efforts, inequality has not abated.

The literature has found conclusive evidence showing that, for example, the United States has increased both economic inequality \citep{piketty2003}, inequality of opportunity \citep{chetty2014} and inequality of achievement \citep{reardon2011}. All of these findings, however, are from the United States. Other countries such as the United Kingdom or other developed countries have their fair share of studies when there is data available. However, very rarely can we compare levels of inequality (of any kind), across several countries with a great degree of confidence.

This paper looks to study the cognitive achievement gap between the 90th and 10th percentile of the SES distribution. This gap can be termed as the SES achievement gap between the top and bottom SES groups. Using the PISA database, we calculate the achievement gap for every country and for all six available PISA waves, building a country panel that shows a time trend of 15 years for each country.

More concretely, we aim to study the evolution of the achievement gap. We aim to answer two broad question: Which countries are changing over time while others are stagnated? Equally important, what explains these changes? Building on the literature on educational institutions we suggest the degree of tracking of a country and it's level of vocational enrollments as possible explanations for the changes of a country over time.

Findings suggest that there's reasonable variation in the evolution of the gap. For example, the United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%. This reduction is also present in other countries such as Germany and Poland. Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap. Robustness checks shows that also the 80/20 and 70/30 replicate the results for nearly all countries.

We then assess whether this gap is correlated with the country's average performance, and find a very strong negative correlation for all country pooled. This suggests that countries with high levels of achievement have low levels of inequality. Yet these results are aggregated for every year. A more disaggregate analysis shows that this pattern is also present for every country over time, although some exceptions are encountered, like the United States where achievement and inequality as positively correlated.

To finish, we find that the degree of tracking of a country seems to be a likely explanation to the changes in the achievement gap, but it should not be studied without factoring in the degree of vocational enrollment. We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap. Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly. Simulations show that if every country switched to little or no tracking their gaps would shrink by about 20\%.

These findings have two specific implications. First, inequality should not be 'generalized' or 'summarized' with one single indicator for many countries. There are widespread differences between countries, where each one should be studied in context. Secondly, now that we know that, further research should attempt to explain why some countries follow such different evolutions. We propose a possible explanation, the degree of tracking and vocational enrollment of the country. Tracking seems to exacerbate inequality, but if coped with a reasonable level of vocational enrollment it can alleviate the negative effects of tracking. Further research should confirm these results hold for other empirical tests and if so, policymakers should incorporate vocational enrollment not only as a 'labor market' solution to youth unemployment but as a compensatory mechanism for highly tracked curriculums.

\section{Literature Review}
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}. But fast-forward until today, we've understood the relationship much better.

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it. For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. To explain these, and other findings, \citet{cunha2006} suggest that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\). Although this sounds logical, it's importance is not understood entirely. The model implies that investments on children are not perfect substitutes at different time points but rather complements. There are specific periods of skill formation where it is much more cost-effective to invest in, which are the early stages of the life course. In fact, when tested against the data, \citet{cunha2006} show that their explanation fits the results of most of the empirical literature very well.

For this reason, children who receive less (or worse) investments have a cumulative disadvantage against children who do receive quality investments. This leads, as early as preschool, to an unequal start in life. That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns \citep{elango15}. Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills. Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}. Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.

This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities. We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

Despite these short-comings, we have indirect evidence of this coming from the vast literature on social mobility. Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitude of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries compared to other European countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}. The important finding came when research emerged the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities. The education system is completely subsidized for all children, otherwise giving opportunities to families who would not have been able to pay. In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}. Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}. These two traits make the Danish system effective in promoting equality relative to the average European country. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable \citep{heckman_oecd}. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
% widening over the years? Gosta, page 3
% income achievement gap, gosta page 3

\citet{reardon2011} finds that the increase in the gaps has occurred predominantly from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by 40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%. Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the 90/10 income gap in the U.S is at about 1.25 standard deviations in the year 2001. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of these estimates. Not necessarily to gauge their statistical significance, but to simply assess how much we can trust their accuracy. It could be that the gap is at 1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. \citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children. This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.

In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend. This follow-up study concentrated solely on kindergarten children for the years 1998, 2006 and 2010. They found that the 90th/10th income gap in readiness closed modestly. Furthermore, using data from fall and spring in the same kindergarten year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010. They also calculated the same changes for a number of personality traits such as self-control and externalizing behavior and found similar results. In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematically increasing at a rate of 0.02, something reasonably close to the previous estimates. Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap). It should be noted that perhaps the reversal of the trend in \citet{reardon2011} would had been evident if data were available for years after 2000, the time-point from which \citet{reardon_portilla} find the reversal.

The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely. They incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.

Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends. The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}. The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada. They find that once the achievement gap is present in early school entry, it does not seem to narrow or widen very much over the life course. In fact, they calculate that the quality of early education can only explain about  30-40\% of the high school SES gap. This suggests that once the achievement gap is present before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous Randomized Controlled Trials (RCT) show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}. One exception is the United Kingdom, which they found to be a country that helps close the gap in early primary years. This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and income support.

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the average European country. A thorough review by \citet{werfhorst_mijs} sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. \citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. The results suggest that early tracking increases educational inequality. While less clear, there is also a tendency for early tracking to reduce mean performance. \citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

Another limitation of the work of \citet{bradbury2015} is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared. They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.  But we should be careful at interpreting these findings causally and should be taken as suggestive at best. For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys. However it should be noted that these studies tackle a completely different question from the above, namely to study cross-sectional differences between many countries, instead of over-time analysis of student gaps. Nonetheless, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap. In other words, does the achievement gap differ between countries? Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. In comparison to the literature on achievement gaps, they find that the U.S has a gap of 1.20 SD in 2001 which increase to around 1.30 in the year 2006 while Germany has a decreasing gap from around 1.25 to 1 standard deviation in the same year-span. However, these numbers vary a lot and carry a great deal of overlapping uncertainty.

They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things. The correlations are very suggestive  as explanatory mechanisms but they are very cautious in drawing causality, specially because they are all reasonably correlated among each other.

One interesting question that is still missing from the literature is whether these country gaps have evolved over time. With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points. That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.

\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from 1964 to 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap. However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase. This is revealing because it does not really pay off to look at a general average once each country has their own distinctive gap and evolution. This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.

One limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies. Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime. However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, these modeling strategies are masking age-specific achievement gaps by controlling for age. We clearly see in \citet{reardon_portilla} that there are age-specific gaps, and they do change at a fast pace in very little time.

In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children. The explanation, although very debated, has been gaining much support in recent years. For countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}. Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}. Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exacerbate the achievement gap.

@@

To investigate the above mentioned questions we will be using the Programme for International Student Assessment (PISA). PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students. Currently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016. This paper takes advantage of these six waves to build a country pseudo-panel where each country-wave will have its corresponding achievement gap, making it possible to study changes in nearly 15 years for each country.

To identify a family's socio-economic status PISA collects several variables that measure different dimensions. Classically, they ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}. This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have small differences. In spite of this, another limitation is the fact that parent's education is measured using the ISCED classification, something that has changed over time. For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.

Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI). This variable attempts to capture the social status of the family, without asking for income information. This index variable was developed by \citet{ganzeboom1996} and later refined by \citep{ganzeboom2010} and it attempts to measure occupational status using a continuous measure. This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}. It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measures the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.

The ESCS was derived from a principal component analysis of standardized variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status. All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only. However, for the purpose of reporting the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries \citep{pisa_2015_results}.

To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indexes can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled indexes for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

Aside from SES, the other relevant variables are test scores for mathematics and literacy. PISA does not provide a single test result for each respondent. Instead, it provides a \emph{series} of 'plausible values' that the child could actually score. As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).

A more intuitive explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\). Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test. This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\). The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}. The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution. In practice each student has 5 scores for each test, that resemble their distribution. Those values are continuous, ranging from 0 to 500, with a mean of 250.

\subsection{Data analysis}

The aim of this paper is to identify, disaggregate and explain country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the difference between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. \citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.

\citet{reardon_portilla}, \citet{anna2016} and \citet{anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable. The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 95th, 50th and 5th \citep{micklewright}. We use the conventional 90/10 cutoffs in the literature following the standard set by \citet{reardon2011}.

For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps. The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization). We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..). "Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares. Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}. The result is a SES gap from an ordinal variable of interest.

As mentioned before, PISA does not provide a single achievement indicator. Instead, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.

To standardize the test score we fit a linear model

% Justify why you use the AGE variable
\begin{equation}
Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
\end{equation}

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds. Controlling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights.

We then adjust \begin{math} \hat{\gamma_i} \end{math} by

% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
\begin{equation}
\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
\end{equation}

% # Take residuals of model and divide by rmse. Multiply that by
% # 1 / sqrt(reliability of each survey), which is .y in the loop.
% .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)

Where \begin{math}\hat{\epsilon_i}\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.

This new standardized variable has a mean of zero. Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric across time. However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased. That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves. But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.

% Try plotting this for each country and wave
% Like, every country as a row and 6 columns for each wave.
% \begin{figure}
% \begin{center}
\begin{figure}

{\centering \includegraphics[width=3.5in,height=3in]{figure/variance_pisa-1}

}

\caption[Standard deviation of test scores across all waves]{Standard deviation of test scores across all waves}\label{fig:variance_pisa}
\end{figure}

% \caption{Standard deviation of test scores across all waves}
% \end{center}
% \end{figure}

We plot the standard deviation of the mathematics test score for all waves in figure 1. The plot suggests that it's something we should not be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous interpretations regarding trends in the of the gaps over time \citep{reardon2011}. PISA has tried to make sure the tests are comparable across waves \citep{pisa2012_technical}, but we still have to adjust for this imprecision.

In order to correct the gap estimates for measurement error, we adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by

\begin{equation}
\hat{\gamma_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
\end{equation}

where \begin{math}r\end{math} is the reliability score of the PISA wave \footnote{Other procedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}. The reliability estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency. Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own. At the moment of writing this paper, they were unavailable. For these reasons we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.

We then fit

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

% Try running also a linear model together with the multilevel model and ompcare results
% Brunno suggested this because countries which might have a very different SES estimate
% might be moved more towards the average.

% mod2 <-
%   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
%        data = .x,
%        weights = .x[[weights_var]])

where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\). We implement this separately for each wave and weight by the wave-specific student sample weights.

Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.

\begin{equation}
\begin{split}
\quad \text{High SES}_j = \alpha_j + \beta_j \\
\quad \text{Low SES}_j = \alpha_j \\
\quad \text{SES gap}_j = \text{High SES}_j - \text{Low SES}_j
\end{split}
\end{equation}

>>>>

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{country_rows} \hlkwb{<-}
  \hlkwd{map_dbl}\hlstd{(adapted_year_data, nrow)} \hlopt{%>%}
  \hlkwd{format}\hlstd{(}\hlkwc{big.mark} \hlstd{=} \hlstr{","}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in map\_dbl(adapted\_year\_data, nrow) \%>\% format(big.mark = "{},"{}): could not find function "{}\%>\%"{}}}\end{kframe}
\end{knitrout}
